A Framework for Substructural Type Systems \thanks{James Wood is supported by an EPSRC Studentship. Robert Atkey is supported by EPSRC grant FIXME.}

### 2022-01-26 tex/esop22/bunchedrules.tex
[ edit ] Major edit to the Semantics section

These observations indicate a way to regularise and streamline the
presentation of this system. Instead of treating each premise and the
conclusion as having potentially unrelated typing and usage
constraints, we make use of combinators for combining premises that will
relate their usage and typing contexts to the conclusion by
construction.
This idea comes from the work of \citet{RPKV20}, including the $\dotto$ and
$\wand$ connectives we introduce later.
To handle binders, which introduce variables, we make
use of a combinator that adds a variable with a given $Ann$-annotation
to an ambient context, without having to explicitly mention the parts
of the context that have not changed. This technique is already
present in some paper presentations of type systems, and is used as a
simplifying technique by Allais et al.~\cite{AACMM21}.
To manage how usage annotations are distributed between premises, we use the
separating ($*$) and sharing ($\dot{\times}$) conjunction connectives from
Bunched Implications~\cite{oHP99}.
To handle the \AgdaInductiveConstructor{!I} rule, we will need a
\emph{scaling} modality, $\gr r \cdot {-}$. The semantics of the bunched
connectives we will use are:
  \begin{align*}
    \dot1\,\grR &\coloneqq 1 \\
    (T \dottimes U)\,\grR &\coloneqq T\,\grR \times U\,\grR \\
    % (T \dotto U)\,\grR &\coloneqq T\,\grR \to U\,\grR \\
    I^*\,\grR &\coloneqq \grR \leq \gr0 \\
    (T \sep U)\,\grR &\coloneqq \Sigma \grP,\grQ.~\plr{\grR \leq \grP + \grQ}
                       \times T\,\grP \times U\,\grQ \\
    (\gr r \cdot T)\,\grR &\coloneqq \Sigma \grP.~\plr{\grR \leq \gr r\grP}
                       \times T\,\grP \\
    % (T \wand U)\,\grP &\coloneqq \Pi \grQ,\grR.~\plr{\grR \leq \grP + \grQ}
    %                    \to T\,\grQ \to U\,\grR
  \end{align*}
An important point to note is that bunched combinators induce
\emph{linear} combinations of substructures, in the sense of the
linear algebra of posemirings described in the previous section.

>>>>

These observations indicate a way to regularise and streamline the
presentation of this system. Instead of treating each premise and the
conclusion as having potentially unrelated typing and usage
constraints, we make use of combinators for combining premises that will
relate their usage and typing contexts to the conclusion by
construction.
This idea comes from the work of \citet{RPKV20}, including the $\dotto$ and
$\wand$ connectives we introduce later.
To handle binders, which introduce variables, we make
use of a combinator that adds a variable with a given $Ann$-annotation
to an ambient context, without having to explicitly mention the parts
of the context that have not changed. This technique is already
present in some paper presentations of type systems, and is used as a
simplifying technique by Allais et al.~\cite{AACMM21}.
To manage how usage annotations are distributed between premises, we use the
separating ($*$) and sharing ($\dot{\times}$) conjunction connectives from
Bunched Implications~\cite{oHP99}.
To handle the \AgdaInductiveConstructor{!I} rule, we will need a
\emph{scaling} modality, $\gr r \cdot {-}$. The semantics of the bunched
connectives we will use in this paper are:
  \begin{align*}
    \dot1\,\grR &\coloneqq 1 \\
    (T \dottimes U)\,\grR &\coloneqq T\,\grR \times U\,\grR \\
    (T \dotto U)\,\grR &\coloneqq T\,\grR \to U\,\grR \\
    I^*\,\grR &\coloneqq \grR \leq \gr0 \\
    (T \sep U)\,\grR &\coloneqq \Sigma \grP,\grQ.~\plr{\grR \leq \grP + \grQ}
                       \times T\,\grP \times U\,\grQ \\
    (\gr r \cdot T)\,\grR &\coloneqq \Sigma \grP.~\plr{\grR \leq \gr r\grP}
                       \times T\,\grP \\
    (T \wand U)\,\grP &\coloneqq \Pi \grQ,\grR.~\plr{\grR \leq \grP + \grQ}
                        \to T\,\grQ \to U\,\grR
  \end{align*}
where the function connectives $\dotto$ and $\wand$ are not used in typing rules, but are used in the rest of the framework.
An important point to note is that bunched combinators induce
\emph{linear} combinations of substructures, in the sense of the
linear algebra of posemirings described in the previous section.

### 2022-01-26 tex/esop22/env.tex
[ fix ] references
[ edit ] Major edit to the Semantics section

We have now seen how to generically build data types of terms for a
given \AgdaRecord{System}, giving us a setting in which to construct
intrinsically well typed and well annotated terms. Before defining a
generic assignment of semantics for terms, we need to develop the
notion of \emph{environment}, a mapping of variables to their semantic
interpretations that is preserved by structural decompositions and
variable bindings.

Given a semantic notion of variable $\V$ : \AgdaFunction{OpenFam}, we
use the notation $\Gamma \sdtstile{}\V A \coloneqq \V\,\Gamma\,A$ for
the type of inhabitants of $\V$ in the context $\Gamma$ at type
$A$. In the non-substructural systems of Allais et al. \cite{AACMM21},
a $\V$-environment $\Gamma \env\V \Delta$ is nothing more than a
function $\forall A \to \Delta \sqni A \to \Gamma \sdtstile{}\V A$,
mapping variables to $\V$-things. In our usage annotated setting
though, we must correctly distribute resources tracked by the
annotations; making sure that we have enough resources in $\Gamma$ to
cover all the demands in $\Delta$. Following the work of \citet{WA20},
this accounting is expressed via the presence of a linear
transformation:

@@

\begin{example}
  Assume $\Ann$ is the natural numbers with ordering given by $=$ and the usual
  addition and multiplication.
  There is a $\sqni$-environment (renaming)
  \[
    \plr{\gr6a : A, \gr0b : B, \gr1c : C, \gr0d : D} \env\sqni
    \plr{\gr1C, \gr2A, \gr4A}.
  \]
  Linear map $\gr\Psi$ is given by the matrix
  \(
    \begin{pmatrix}
      \gr0 & \gr0 & \gr1 & \gr0 \\
      \gr1 & \gr0 & \gr0 & \gr0 \\
      \gr1 & \gr0 & \gr0 & \gr0
    \end{pmatrix}
  \),
  which we can check satisfies the required inequality.
  The values are given by
  \begin{align*}
    \gr0a : A, \gr0b : B, \gr1c : C, \gr0 : D &\sqni c : C \\
    \gr1a : A, \gr0b : B, \gr0c : C, \gr0 : D &\sqni a : A \\
    \gr1a : A, \gr0b : B, \gr0c : C, \gr0 : D &\sqni a : A.
  \end{align*}

  We can read from the columns of the matrix what happened to each of the
  variables in $\Gamma$.
  The first column, corresponding to variable $\gr6a : A$, contains two $\gr1$s
  because it has been duplicated (via contraction).
  Meanwhile, the second and fourth columns are all $\gr0$ because variables
  $b$ and $d$ have been discarded (via weakening).
  The third column contains one $\gr1$ because $c$ is used once.
  This $\gr1$ appears above the $\gr1$s to its left because $c$ has been
  permuted (via exchange) past $a$.
  Each of the rows in the matrix is a basis vector because variables can only
  be formed in contexts with basis annotations or less.
\end{example}

@@

\begin{lemma}[\AgdaFunction{relocate}]\label{thm:env-resize}
  Given an environment $\rho : \grP\gamma \env\V \grQ\delta$ and a $\grPprime$
  and a $\grQprime$ such that $\grPprime \leq \grQprime\plr{\rho.\gr\Psi}$,
  there is also an environment of type $\grPprime\gamma \env\V \grQprime\delta$
  with the same linear map and action on variables.
\end{lemma}
% \begin{proof}
%   The only part of the definition of an environment dependent on $\grP$ or
%   $\grQ$ is the constraint $\grP \leq \grQ\gr\Psi$, which we are able to
%   replace for $\grPprime$ and $\grQprime$.
% \end{proof}
Relocation will be used when pushing environments into subterms in
\secref{XX}.

@@

>>>>

We have now seen how to build data types of intrinsically well typed
and well uasged terms for a given \AgdaRecord{System}. In the next
section, we will define a generic traversal function that assigns a
``semantics'' to each term. Traversals operate on open terms, so they
need a way to assign semantics to variables in a typed and usage
respecting manner. This is the function fulfiled by
\emph{environments}.

Given a semantic notion of variable $\V$ : \AgdaFunction{OpenFam}, we
use the notation $\Gamma \sdtstile{}\V A$ meaning $\V\,\Gamma\,A$ for
the type of inhabitants of $\V$ in the context $\Gamma$ at type
$A$. In the non-substructural systems of Allais et al. \cite{AACMM21},
a $\V$-environment $\Gamma \env\V \Delta$ is nothing more than a
function $\forall A \to \Delta \sqni A \to \Gamma \sdtstile{}\V A$,
mapping variables to $\V$-things. In our usage annotated setting
though, we must correctly distribute resources tracked by the
annotations; making sure that we have enough resources in $\Gamma$ to
cover all the demands in $\Delta$. Following the work of \citet{WA20},
this accounting is expressed via the presence of a linear
transformation:

@@

\begin{example}
  Assume $\Ann$ is the natural numbers with ordering given by $=$ and the usual
  addition and multiplication.
  There is a $\sqni$-environment (a renaming)
  \[
    \plr{\gr6a : A, \gr0b : B, \gr1c : C, \gr0d : D} \env\sqni
    \plr{\gr1C, \gr2A, \gr4A}.
  \]
  The mapping of variables to variables and matrix giving the linear
  map $\gr\Phi$ are:
  \begin{displaymath}
    \begin{array}{ll}
    \gr0a : A, \gr0b : B, \gr1c : C, \gr0 : D &\sqni c : C \\
    \gr1a : A, \gr0b : B, \gr0c : C, \gr0 : D &\sqni a : A \\
    \gr1a : A, \gr0b : B, \gr0c : C, \gr0 : D &\sqni a : A
    \end{array}
    \qquad\qquad
    \begin{pmatrix}
      \gr0 & \gr0 & \gr1 & \gr0 \\
      \gr1 & \gr0 & \gr0 & \gr0 \\
      \gr1 & \gr0 & \gr0 & \gr0
    \end{pmatrix}
  \end{displaymath}
  Note that $(\gr 6\,\gr 0\,\gr 1\,\gr 0) = (\gr 1\,\gr 2\,\gr 4)\gr \Phi$.
  % We can read from the columns of the matrix what happened to each of the
  % variables in $\Gamma$.
  The first column of $\gr\Phi$, corresponding to variable $\gr6a : A$, contains two $\gr1$s
  because it has been duplicated (via contraction).
  The second and fourth columns are all $\gr0$ because variables
  $b$ and $d$ have been discarded (via weakening).
  The third column contains one $\gr1$ because $c$ is used once.
  This $\gr1$ appears above the $\gr1$s to its left because $c$ has been
  permuted (via exchange) past $a$.
  Each of the rows in the matrix is a basis vector because variables can only
  be formed in contexts with basis annotations or less.
\end{example}

@@

\begin{lemma}[\AgdaFunction{relocate}]\label{thm:env-resize}
  Given an environment $\rho : \grP\gamma \env\V \grQ\delta$ and a $\grPprime$
  and a $\grQprime$ such that $\grPprime \leq \grQprime\plr{\rho.\gr\Psi}$,
  there is also an environment of type $\grPprime\gamma \env\V \grQprime\delta$
  with the same linear map and action on variables.
\end{lemma}
% \begin{proof}
%   The only part of the definition of an environment dependent on $\grP$ or
%   $\grQ$ is the constraint $\grP \leq \grQ\gr\Psi$, which we are able to
%   replace for $\grPprime$ and $\grQprime$.
% \end{proof}
Relocation will be used when pushing environments into subterms in
\secref{sec:traversal}.

@@

\subsection{Renameability}

\todo{Sort this out: we get a category of renamings, and the comonad a la AACMM}
We saw in \cref{sec:env} that our notion of environment specialises to
simultaneous renaming by setting $\V$ to be $\sqni$.
This is encoded below as
\AgdaFunction{\AgdaUnderscore{}$\Rightarrow^r$\AgdaUnderscore{}}.

\ExecuteMetaData[\Renamingtex]{Ren}

In a setting where new variables can be bound in the middle of a derivation,
it is important that the values we carry around while traversing a term can
handle the existence of variables that appear but they do not use.
We call any such notion of value \emph{renameable}.
The cofree renameable \AgdaFunction{OpenType} on an \AgdaFunction{OpenType}
\AgdaBound{T} is \AgdaFunction{$\Box^r$}\AgdaSpace{}\AgdaBound{T}, with
\AgdaBound{T} then being \AgdaFunction{Renameable} if it forms a
\AgdaFunction{$\Box^r$}-coalgebra.

\ExecuteMetaData[\Renamingtex]{Boxr}
\ExecuteMetaData[\Renamingtex]{Renameable}

A renameable notion of value gives rise to a renameable notion of environment,
essentially by renaming each contained value in an appropriate way.

\begin{lemma}[\AgdaFunction{ren\textasciicircum{}Env}]\label{thm:env-ren}
  If $\plr{-} \sdtstile{}\V A$ is renameable for all $A$, then so is
  $\plr{-} \env\V \Delta$ for all $\Delta$.
\end{lemma}
\bob{Proof probably isn't needed}
\begin{proof}
  For renameable $\V$, we can implement $\mathrm{lift}$ as in
  \cref{thm:env-comp} with $\U \coloneqq {\sqni}$ and
  $\V \coloneqq \W \coloneqq \V$.
  Renaming the environment then becomes an instance of environment composition.
\end{proof}

### 2022-01-26 tex/esop22/semantics.tex
[ edit ] Major edit to the Semantics section

Our goal in this section is to define \AgdaFunction{semantics}, a
recursor that turns a term into a \AgdaBound{$\C$}-value using a
\AgdaBound{$\V$}-environment, in a type preserving way:\bob{Get rid of
  ``body'' here}

The \AgdaBound{$\V$} and \AgdaBound{$\C$} are \AgdaFunction{OpenFam}s,
representing the interpretations of variables and terms
respectively. In \cref{sec:traversal} we will see the data that must
be provided to make a \AgdaFunction{semantics} for a given
system. Before that, we must see how to deal with the two complicated
features of our syntax: the usage annotations (\cref{sec:functorial})
and variable binding (\cref{sec:kripke}). \todo{fwd ref to where these are used}

@@

A basic property of the universe of syntaxes we described in \cref{sec:syntax}
is that every syntax supports a functorial action on subterms, realised by the function \AgdaFunction{map-s}.
Its says that to map a function \AgdaBound{f}
over a layer of syntax, there must be a linear map \AgdaBound{F} relating the
domain and codomain usage contexts, and \AgdaBound{f} should be usable
wherever the domain and codomain usage contexts are similarly related by
\AgdaBound{F}.

This generality is needed because usage contexts change between
a term and its immediate subterms, but there will be a linear combination of
the subterms' usage contexts for the subterms that make the original usage contexts.
\AgdaBound{X} and \AgdaBound{Y} are \AgdaFunction{ExtOpenFam}s, with
\AgdaBound{$\Theta$} being the context extension for a subterm (i.e., the
variables newly bound in that subterm).
Unlike usage annotations, types are preserved throughout a term (including
\AgdaBound{$\gamma$}, \AgdaBound{$\delta$}, and the implicit type of the term).

@@

The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
works through the structure \AgdaBound{ps} of premises of the rule applied,
acting on each subterm it finds.
Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
--- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
and \AgdaBound{Q$'$}.
In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
Linearity of \AgdaBound{F} yields $\grQ_M$ and $\grQ_N$ such that
$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaBound{map-p} recursively at
$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
The cases for \AgdaInductiveConstructor{`$\cdot$} and
\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
of linearity.
In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
\AgdaInductiveConstructor{`$\dot\times$}, which exist in fully structural
systems, do not involve any changes in the usage contexts.

@@

In \secref{sec:terms}, we defined \AgdaFunction{Scope} to let a
judgement-indexed family admit context extensions. However, a key
component of our generic semantic traversal is to make use of the open
family \AgdaBound{$\V$} of \emph{values}, which are the sort of thing
we store in an environment.  The definition \AgdaFunction{Kripke}
gives an alternative to \AgdaFunction{Scope} which interprets the
newly bound variables via a requirement of $\V$-values rather than
extra assumptions for the $\C$-computation.

@@

Just as $\dotto$ is the internal hom for $\dottimes$, we introduce $\wand$ as
the internal hom for $\sep$, with \AgdaFunction{$\wand^c$} acting on open types
analogously to \AgdaFunction{$\sep^c$}.
It is given by $\plr{T \wand U}\,\grP \coloneqq
\Pi\grQ,\grR.~\plr{\grR \leq \grP + \grQ} \to T\,\grQ \to U\,\grR$.

For understanding the meaning of the type family, \AgdaFunction{Wrap} can be
ignored.
\AgdaFunction{Wrap} is a device that turns any type family into an equivalent
type family that is judgementally injective in its indices.
It turns the type family into a parametrised record with a single field
\AgdaField{get} whose type is the type in the body of the $\lambda$-abstraction.
If $\Delta$ is of the form $\gr{s_1}B_1, \ldots, \gr{s_n}B_n$, then
\ExecuteMetaData[\Snippetstex]{KripkeVCDGA}\ is equivalent to
\ExecuteMetaData[\Snippetstex]{KripkeExpanded}\ by Currying.
That is to say, the Kripke function is expecting a value for each newly bound
variable, at the multiplicity of its annotation, together with the resources
supporting each of those values.
The need for \AgdaFunction{$\Box^r$} is exemplified by the following construct
\AgdaFunction{reify}, where we weaken \AgdaBound{$\Gamma$} by a $\gr0$ed-out
version of \AgdaBound{$\Delta$}.
The \AgdaBound{$\Delta$} then gets filled in by the $\V$-values.

\bob{Move this para} At this point we introduce a minor generalisation to \AgdaFunction{OpenFam}
and \AgdaFunction{ExtOpenFam}:
\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam} and
\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---ExtOpenFam}.
We obtain the definition of \AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam}
by replacing the textual occurrence of \AgdaBound{Ty} by the parameter
\AgdaBound{I}.
This means that \AgdaBound{A} in the definition of \AgdaFunction{Kripke} has
type \AgdaBound{I}, rather than specifically \AgdaBound{Ty}.
We use this generality later in \AgdaFunction{extend}, setting \AgdaBound{I}
to \AgdaDatatype{Ctx}.

\subsection{Renameability}

\todo{Maybe this goes in the Environment section.}
We saw in \cref{sec:env} that our notion of environment specialises to
simultaneous renaming by setting $\V$ to be $\sqni$.
This is encoded below as
\AgdaFunction{\AgdaUnderscore{}$\Rightarrow^r$\AgdaUnderscore{}}.

\ExecuteMetaData[\Renamingtex]{Ren}

In a setting where new variables can be bound in the middle of a derivation,
it is important that the values we carry around while traversing a term can
handle the existence of variables that appear but they do not use.
We call any such notion of value \emph{renameable}.
The cofree renameable \AgdaFunction{OpenType} on an \AgdaFunction{OpenType}
\AgdaBound{T} is \AgdaFunction{$\Box^r$}\AgdaSpace{}\AgdaBound{T}, with
\AgdaBound{T} then being \AgdaFunction{Renameable} if it forms a
\AgdaFunction{$\Box^r$}-coalgebra.

\ExecuteMetaData[\Renamingtex]{Boxr}
\ExecuteMetaData[\Renamingtex]{Renameable}

A renameable notion of value gives rise to a renameable notion of environment,
essentially by renaming each contained value in an appropriate way.

\begin{lemma}[\AgdaFunction{ren\textasciicircum{}Env}]\label{thm:env-ren}
  If $\plr{-} \sdtstile{}\V A$ is renameable for all $A$, then so is
  $\plr{-} \env\V \Delta$ for all $\Delta$.
\end{lemma}
\bob{Proof probably isn't needed}
\begin{proof}
  For renameable $\V$, we can implement $\mathrm{lift}$ as in
  \cref{thm:env-comp} with $\U \coloneqq {\sqni}$ and
  $\V \coloneqq \W \coloneqq \V$.
  Renaming the environment then becomes an instance of environment composition.
\end{proof}

@@

The aim of this subsection is to give an alternative recursion principle for
terms that incorporates some of the environment-handling seen in the
implementations of renaming and substitution.
The rest of this section assumes the following: a renameable open family
\AgdaBound{$\V$} that embeds into the open family \AgdaBound{$\C$}, and an
algebra for a layer of syntax at \AgdaBound{$\C$}.

@@

Under the assumption that \AgdaBound{$\V$} is renameable, we can decompose
\cref{thm:lr-bind} as
\AgdaFunction{reify}\AgdaSpace{}\AgdaOperator{\AgdaFunction{$\circ$}}%
\AgdaSpace{}\AgdaFunction{extend}, with \AgdaFunction{extend} defined below.
We can think of \AgdaFunction{extend} as our best effort to extend an
environment by \AgdaBound{$\Theta$} without access to an identity environment
at \AgdaBound{$\Theta$}.
We instead take \AgdaBound{$\rho$}, which is an environment covering
\AgdaBound{$\Delta$}, and \AgdaBound{$\sigma$}, which is an environment
covering \AgdaBound{$\Theta$}, and glue them together using
\cref{thm:env-rules}, after having renamed \AgdaBound{$\rho$} via
\cref{thm:env-ren} to make it fit the new context \AgdaBound{$\Gamma^+$}
(intended to be \ExecuteMetaData[\Snippetstex]{GT}).

@@

For \AgdaFunction{body}, we are given a subterm \AgdaFunction{M}, to which we
want to apply \AgdaFunction{semantics}.
To do so, we need an extended version of the initial environment
\AgdaBound{$\rho$}.
The best we can achieve without identity environments for \AgdaBound{$\V$} is
a Kripke function returning an extended environment.
We use \AgdaFunction{mapK$\C$} to post-compose this function by the
\AgdaSymbol{$\lambda$}-function taking an extended environment and acting with
it on \AgdaBound{M}.

@@

>>>>

Given a $\V$-environment $\Gamma \Rightarrow \Delta$, the function
\AgdaFunction{semantics} we define in this section assigns a
$\C$-value in context $\Gamma$ to every term in context $\Delta$,
where $\C$ is an \AgdaFunction{OpenFam} being the carrier of the
semantic interpretation of terms ($\V$ being the semantic
interpretation of variables). Before we can define
\AgdaFunction{semantics}, we need to treat recursion through rules'
premises (\cref{sec:functorial}) and extension of environments when
going under variable binders (\cref{sec:kripke}).

% Our goal in this section is to define \AgdaFunction{semantics}, a
% recursor that turns a term into a \AgdaBound{$\C$}-value using a
% \AgdaBound{$\V$}-environment, in a type preserving way:\bob{Get rid of
%   ``body'' here}

% \ExecuteMetaData[\Semanticstex]{semantics-type}

% The \AgdaBound{$\V$} and \AgdaBound{$\C$} are \AgdaFunction{OpenFam}s,
% representing the interpretations of variables and terms
% respectively. In \cref{sec:traversal} we will see the data that must
% be provided to make a \AgdaFunction{semantics} for a given
% system. Before that, we must see how to deal with the two complicated
% features of our syntax: the usage annotations (\cref{sec:functorial})
% and variable binding (\cref{sec:kripke}). \todo{fwd ref to where these are used}

@@

A basic property of the universe of syntaxes we described in \cref{sec:syntax}
is that every syntax supports a functorial action on subterms, realised by the function \AgdaFunction{map-s}.
Its type says that to map a function \AgdaBound{f}
over a layer of syntax, there must be a linear map \AgdaBound{F} relating the
domain and codomain usage contexts, and \AgdaBound{f} should be usable
wherever the domain and codomain usage contexts are similarly related by
\AgdaBound{F}.

@@

This generality is needed because usage contexts change between
a term and its immediate subterms---they are decomposed according to the bunched connectives used in the rules.
\AgdaBound{X} and \AgdaBound{Y} are \AgdaFunction{ExtOpenFam}s, with
\AgdaBound{$\Theta$} being the context extension for a subterm (i.e., the
variables newly bound in that subterm).
Unlike usage annotations, types in the contexts \AgdaBound{$\gamma$} and \AgdaBound{$\delta$}, and the conclusion types implicit here, are preserved throughout.
This is the essence of the usage annotation based approach---we use traditional techniques for variable binding, with the usage annotations layered on top.

The heart of \AgdaFunction{map-s} is \AgdaFunction{map-p}, which recursively
works through the structure \AgdaBound{ps} of premises of the rule applied,
acting on each subterm it finds.
Here, particularly in the clauses for \AgdaInductiveConstructor{`$\sep$} and
\AgdaInductiveConstructor{`$\cdot$}, we see why it is not enough for the
function on subterms to apply at usage contexts \AgdaBound{P} and \AgdaBound{Q}
--- rather, it also needs to apply at any similarly related \AgdaBound{P$'$}
and \AgdaBound{Q$'$}.
In the case of \AgdaInductiveConstructor{`$\sep$}, we have that
$\grP \leq \grP_M + \grP_N$, with \AgdaBound{M} and \AgdaBound{N} being
collections of subterms in usage contexts $\grP_M$ and $\grP_N$, respectively.
Linearity of \AgdaBound{F} yields $\grQ_M$ and $\grQ_N$ such that
$\grQ \leq \grQ_M + \grQ_N$ and we use \AgdaBound{map-p} recursively at
$(\grP_M, \grQ_M)$ and $(\grP_N, \grQ_N)$ on \AgdaBound{M} and \AgdaBound{N}.
The cases for \AgdaInductiveConstructor{`$\cdot$} and
\AgdaInductiveConstructor{`$I^*$} are similar, each using a different aspect
of linearity.
In contrast, the cases for \AgdaInductiveConstructor{`$\dot1$} and
\AgdaInductiveConstructor{`$\dot\times$}, which are the only constructors used in fully structural
systems, do not involve any changes in the usage contexts.

@@

At this point we introduce a minor generalisation to
\AgdaFunction{OpenFam} and \AgdaFunction{ExtOpenFam}:
\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam} and
\AgdaBound{I}\AgdaSpace{}\AgdaFunction{---ExtOpenFam}.  We obtain the
definition of \AgdaBound{I}\AgdaSpace{}\AgdaFunction{---OpenFam} by
replacing the textual occurrence of \AgdaBound{Ty} by the parameter
\AgdaBound{I}.

The definition \AgdaFunction{Kripke}\,$\V$\,$\C$\,$\Delta$ is a kind
of function space that describes a $\C$ value parameterised by
$\Delta$-many additional $\V$s (all correctly typed and usage
annotated). It is used to describe how to go under binders in a
Higher-Order Abstract Syntax style---to go under a binder we must
provide semantic interpretations for all the additional variables:

% When going under binders during a recursion, the context will be extended by some $\Theta$. This means that the current environemnt must be extended with $\Theta$s-worth of $\V$s

% we need the ability to say that

% Kripke V C is given the extension \Theta

% In \secref{sec:terms}, we defined \AgdaFunction{Scope} to let a
% judgement-indexed family admit context extensions. However, a key
% component of our generic semantic traversal is to make use of the open
% family \AgdaBound{$\V$} of \emph{values}, which are the sort of thing
% we store in an environment.  The definition \AgdaFunction{Kripke}
% gives an alternative to \AgdaFunction{Scope} which interprets the
% newly bound variables via a requirement of $\V$-values rather than
% extra assumptions for the $\C$-computation.

@@

\AgdaFunction{Wrap}
is a device that turns any type family into an equivalent type family
that is judgementally injective in its indices, which helps with
Agda's type inference.
It turns the type family into a parametrised
record with a single field \AgdaField{get} whose type is the type in
the body of the $\lambda$-abstraction.
For understanding the meaning of
\AgdaFunction{Kripke}, \AgdaFunction{Wrap} can be ignored.

If $\Delta$ is of the form $\gr{s_1}B_1, \ldots, \gr{s_n}B_n$, then
\ExecuteMetaData[\Snippetstex]{KripkeVCDGA}\ is equivalent to
\ExecuteMetaData[\Snippetstex]{KripkeExpanded}\ by Currying.  That is
to say, the Kripke function is expecting a value for each newly bound
variable, at the multiplicity of its annotation, together with the
resources supporting each of those values. We use the ``magic wand''
function space here to enforce the invariant that the freshly bound
variables have usage annotations that are added to the existing
variables, not shared with them. The use of the
\AgdaFunction{$\Box^r$} modality ensures that we can still use it in
the presence of additional variables introduced by weakening.

\AgdaFunction{Kripke} is functorial in the $\C$ argument, as witnessed
by the \AgdaFunction{mapKC} function, which is essentially
post-composition:

\ExecuteMetaData[\Semanticstex]{mapKC}

% is exemplified by the following construct
% \AgdaFunction{reify}, where we weaken \AgdaBound{$\Gamma$} by a $\gr0$ed-out
% version of \AgdaBound{$\Delta$}.
% The \AgdaBound{$\Delta$} then gets filled in by the $\V$-values.

% \bob{Move this para}
% This means that \AgdaBound{A} in the definition of \AgdaFunction{Kripke} has
% type \AgdaBound{I}, rather than specifically \AgdaBound{Ty}.
% We use this generality later in \AgdaFunction{extend}, setting \AgdaBound{I}
% to \AgdaDatatype{Ctx}.

@@

We can now state the data required to implement a traversal assigning
semantics to terms. For open families $\V$ and $\C$, interpreting
variables and terms respectively, we assume that $\V$ is renamable,
that $\V$ is embeddable in $\C$, and that we have an algebra for a
layer of syntax, where bound variables are handled using the Kripke
function space:

% The aim of this subsection is to give an alternative recursion principle for
% terms that incorporates some of the environment-handling seen in the
% implementations of renaming and substitution.
% The rest of this section assumes the following: a renameable open family
% \AgdaBound{$\V$} that embeds into the open family \AgdaBound{$\C$}, and an
% algebra for a layer of syntax at \AgdaBound{$\C$}.

@@

@@

For \AgdaFunction{body}, we are given a subterm \AgdaBound{M}, to
which we want to apply \AgdaFunction{semantics}.  To do so, we need an
extended version of the initial environment \AgdaBound{$\rho$}. We
express this as the generation of a Kripke function that produces the
extended enviroment given interpretations of the fresh variables. We
take \AgdaBound{$\rho$}, which is an environment covering
\AgdaBound{$\Delta$}, and \AgdaBound{$\sigma$}, which is an
environment covering \AgdaBound{$\Theta$}, and glue them together
using the inductive rules for generating environments, after having
renamed \AgdaBound{$\rho$} via \cref{thm:env-ren} to make it fit the
new context \AgdaBound{$\Gamma^+$} (intended to be
\ExecuteMetaData[\Snippetstex]{GT}):

% The best we can achieve without identity environments for \AgdaBound{$\V$} is
% a Kripke function returning an extended environment.
To define \AgdaFunction{body}, we use \AgdaFunction{mapK$\C$} to
post-compose the environment extension by the
\AgdaSymbol{$\lambda$}-function taking an extended environment and
acting with it on \AgdaBound{M}.

@@

% \todo{FIX} Under the assumption that \AgdaBound{$\V$} is renameable, we can decompose
% \cref{thm:lr-bind} as
% \AgdaFunction{reify}\AgdaSpace{}\AgdaOperator{\AgdaFunction{$\circ$}}%
% \AgdaSpace{}\AgdaFunction{extend}, with \AgdaFunction{extend} defined below.
% We can think of \AgdaFunction{extend} as our best effort to extend an
% environment by \AgdaBound{$\Theta$} without access to an identity environment
% at \AgdaBound{$\Theta$}.
